Supporting Information:\\ Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-05-22 Manuscript/manuscriptPSsuppl.tex
Updates based on comments and text from Tom Piggot sent by email

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Markus Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot}
\affiliation{Southampton, United Kingdom}

@@

\noindent
{\it DOPS and POPS bilayers.} \\
The data in Table \ref{PSsystems} and Refs \citenum{charmm36DOPS303K,charmm36POPS298K,charmm36POPS298Kpotassium}
\todo{T. Piggot, please write simulation details} \\

\noindent
{\it POPC:POPS mixtures without additional ions}\\
Simulations of POPC:POPS (5:1) mixture containing total 132 (110:22) lipids
with potassium and sodium countertions were ran using CHARMM36 force field.
The data are available from Refs. \citenum{charmm36pops+83popcT298Kpiggot}
and \citenum{charmm36pops+83popcT298KpiggotSODIUM}, respectively\todo{T. Piggot, please write simulation details}.

Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids
with potassium countertions were ran using CHARMM36 force field.
The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}. POPC:POPS (both 1:1 and 5:1) mixtures were prepared with 300 lipids in total (150:150 and 250:50, respectively) with the neutralizing ions with the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298 K. For further details see table \ref{mixedIONsystemsMONOVALENT}. 
 \\

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium} \\
POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
were simulated with the additional calcium concentrations.
The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
Systems with 0.15 M and 1 M calcium chloride were prepared using the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298 K. For further details see table S1. \\

\subsection{CHARMM36ua}
\noindent
{\it DOPS and POPS bilayers.} \\
The data in Table \ref{PSsystems} and Refs \citenum{charmm36uaDOPS303K,charmm36uaPOPS298K}
\todo{T. Piggot, please write simulation details} \\

@@

\noindent
{\it POPS bilayers.} \\
The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
\todo{T. Piggot and M. Javanainen, please write simulation details} \\
\hl{Piggot: please write about system with Na$^+$ counterions}.

@@

\subsection{Slipids}

\noindent
{\it POPS}
\todo{POPS \cite{slipidsPOPS298K} to be written by Piggot}

\noindent
{\it DOPS with 128 lipids.}
\todo{DOPS with 128 lipids \cite{slipidsDOPS303K} to be written by Piggot}

@@

\noindent
{\it DOPS and POPS bilayers.} \\
The data in Table \ref{PSsystems} and Refs \citenum{bergerPOPS298K,bergerDOPS303K}
\todo{T. Piggot, please write simulation details} \\

@@

\subsection{GROMOS-CKP and GROMOS-CKPM}
  CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges~\cite{mukhopadhyay04})
   and CKP refers to the version with more Gromos compatible version
   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
\todo{T. Piggot, please write the description of force field.}

\noindent
{\it DOPS and POPS bilayers.} \\
The data in Table \ref{PSsystems} and Refs. \citenum{ckp1POPS303K,ckp2POPS303K,ckp1DOPS303K,ckp2DOPS303K}
\todo{T. Piggot, please write simulation details} \\

\noindent
    {\it POPC:POPS (5:1) mixtures without additional ions}\\
    The data in Table \ref{mixedIONsystems} and Refs. \citenum{POPCpopsGROMOSCKPwithNa,POPCpopsGROMOSCKPMwithNa}
\todo{T. Piggot, please write simulation details} \\

@@

\noindent
{\it Thomas Piggot}

>>>>

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Markus Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas J. Piggot}
\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}

@@

\noindent
{\it DOPS and POPS bilayers.}
%The data in Table \ref{PSsystems} and Refs \citenum{charmm36DOPS303K,charmm36POPS298K,charmm36POPS298Kpotassium}
Starting structures for CHARMM36 DOPS and POPS simulations were constructed using the CHARMM-GUI webserver~\cite{lee16,jo18}.
The systems contained a total of 128 lipids (64 per leaflet of the membrane), 35 waters per lipid, and 128 Na$^+$ ions.
Force field parameters for DOPS and POPS were taken from Venable et al. and included the modified parameters for interactions
with Na$^+$ ions \cite{venable13}. Water was treated using the CHARMM TIP3P model \cite{durell94,eyal96}.
%https://pubs.acs.org/doi/10.1021/j100059a038 and https://aip.scitation.org/doi/10.1063/1.472061].
For POPS, an identical system was also constructed in which the 128 Na$^+$ ions were replaced with 128 K$^+$ ions.

Simulations of these systems were performed for 500 ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
A 2 fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
The simulations were maintained at temperatures of 303 K and 298 K for DOPS and POPS respectively using the Nosé-Hoover thermostat \cite{nose84,hoover85}
with a coupling constant of 1 ps. A pressure of 1 bar was maintained using the Parrinello-Rahman \cite{parrinello81} method with a coupling constant of 5 ps.
Pressure coupling was applied in a semi-isotropic manner with the x and y dimensions, in the plane of the bilayer, fluctuating independently of the z dimension.
Standard CHARMM36 force field methods were applied for the simulation cut-offs: van der Waals interactions were truncated at 1.2 nm with the interactions
switched off between 1.0 nm and 1.2 nm; Coulombic interactions were truncated at 1.2 nm with long-range interactions treating using PME~\cite{darden93,essman95}.

Analyses on these simulations was performed on the final 100 ns of the simulations.

\noindent
{\it POPC:POPS mixtures without additional ions}
Simulations of POPC:POPS (5:1) mixtures containing a total of 132 lipids (110 POPC and 22 POPS) were constructed using the CHARMM-GUI~\cite{lee16,jo18}.
Two identical systems were constructed, apart from one system was neutralised through the addition of 22 Na$^+$ ions while the other system
contained 22 K$^+$ ions. As per the pure POPS CHARMM36 simulations, 35 water per lipid were added. All other system and simulation parameters
and settings were identical to those used in the CHARMM36 POPS simulations.

%Simulations of POPC:POPS (5:1) mixture containing total 132 (110:22) lipids
%with potassium and sodium countertions were ran using CHARMM36 force field.
%The data are available from Refs. \citenum{charmm36pops+83popcT298Kpiggot}
%and \citenum{charmm36pops+83popcT298KpiggotSODIUM}, respectively\todo{T. Piggot, please write simulation details}.

Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids (250:50 and 150:150, respectively)
with neutralizing potassium countertions 
%ran using CHARMM36 force field.
%The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
%POPC:POPS (both 1:1 and 5:1) mixtures
were prepared
%with 300 lipids in total
%with the  ions
using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298 K. For further details see table \ref{mixedIONsystemsMONOVALENT}. 
 \\

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium} \\
POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
%were simulated with the additional calcium concentrations.
%The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
and 0.26 M and 1 M CaCl$_2$ were prepared using the CHARMM-GUI \cite{lee16,jo18}.
Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
by the CHARMM-GUI at 298 K.

\subsection{CHARMM36ua}
\noindent
{\it DOPS and POPS bilayers.} \\
%The data in Table \ref{PSsystems} and Refs \citenum{charmm36uaDOPS303K,charmm36uaPOPS298K}
Starting structures for CHARMM36-UA DOPS and POPS simulations were taken from those generated for the CHARMM36 simulations (see above),
with any extraneous hydrogen atoms removed in the lipid tails. Force field parameters for DOPS and POPS were constructed through a
combination of the published CHARMM36 PS~\cite{venable13} and CHARMM36-UA PC parameters~\cite{lee14}.
% [REF: https://pubs.acs.org/doi/10.1021/jp410344g].
All other simulation parameters and conditions were identical to those described for the CHARMM36 all-atom DOPS and POPS simulations.

@@

\noindent
{\it POPS bilayers.}
%The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
%\todo{T. Piggot and M. Javanainen, please write simulation details} \\
%\hl{Piggot: please write about system with Na$^+$ counterions}.
Force field parameters for POPS simulations with the MacRog OPLS-AA force field~\cite{maciejewski14,kulig15b,rog16}
were constructed manually using the provided PS parameters as a guide~\cite{rog16}. The original parameters were
not used due to an incorrect arrangement of the {\it sn}-1 and {\it sn}-2 tails (i.e. the lipid was OPPS and not POPS).
The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na+ ions) and was converted into
the correct format using a script to adjust the atom order. Water was treated using the TIP3P model~\cite{jorgensen83}.
Simulation parameters were chosen to closely mimic those used in the original force field publications~\cite{maciejewski14,kulig15b,rog16}:
van der Waals interactions were truncated at 1.0 nm with a dispersion correction applied to the energy and pressure;
Coulombic interactions were truncated at 1.0 nm with PME used to treat interactions beyond this cut-off.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
including use of the Verlet cut-off scheme~\cite{Pall13}, apart from the use of LINCS to constrain all bonds during these simulations.

@@

\subsection{Slipids}
%
%\noindent
%{\it POPS}
%\todo{POPS \cite{slipidsPOPS298K} to be written by Piggot}%%
%
%\noindent
%{\it DOPS with 128 lipids.}
%\todo{DOPS with 128 lipids \cite{slipidsDOPS303K} to be written by Piggot}
%
\noindent
{\it POPS and DOPS with 128 lipids}
Starting structures for the Slipids POPS and DOPS (with 128 lipids) simulations were the same as constructed
for the CHARMM36 simulations. Force field parameters were taken from the published Slipids PS model,
however simulation cut-offs used herein were different to these previously published simulations~\cite{jambeck2012another}.
% [REF: https://pubs.acs.org/doi/10.1021/ct300777p].
This was due to the use of the older GROMACS group-based cut-off scheme in these previous simulations.
In the simulations reported in this work, the Verlet cut-off scheme was used. Van der Waals interactions were
truncated at 1.0 nm with LJ-PME~\cite{wennberg15}
% [REF: https://pubs.acs.org/doi/10.1021/acs.jctc.5b00726]
applied to account for the long-range interactions.
Coulombic interactions were truncated at 1.0 nm with PME used for the long-range interactions.
Full testing and validation of these settings with the Slipids PS force field will be provided in an additional forthcoming work.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations.

@@

\noindent
{\it DOPS and POPS bilayers.} \\
%The data in Table \ref{PSsystems} and Refs \citenum{bergerPOPS298K,bergerDOPS303K}
%\todo{T. Piggot, please write simulation details} \\
Starting structures for DOPS and POPS simulations performed with a Berger-based PS force field
were taken from those produced for the CHARMM36 simulations. A script was used to re-order the
atoms and to remove any extraneous hydrogen atoms. Force field parameters were taken from those
published by Mukhopadhyay et al.~\cite{mukhopadhyay04}
% [REF: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1303994/]
and the SPC water model was used~\cite{berendsen81}. The simulation cut-offs closely matched
those of Mukhopadhyay et al.: van der Waals interactions were truncated at 1.0 nm;
Coulombic interactions were truncated at 1.0 nm with PME applied to account for the interactions beyond the cut-off.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
including the use of the Verlet cut-off scheme, apart from all bonds were constrained using LINCS in these simulations.

@@

\subsection{GROMOS-CKP and GROMOS-CKPM}
%  CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
%   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
%   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges~\cite{mukhopadhyay04})
%   and CKP refers to the version with more Gromos compatible version
%   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
%\todo{T. Piggot, please write the description of force field.}
The GROMOS-CKP force field (CKP stands for Chandrasekhar-Kukol-Piggot) contains a set of GROMOS compatible lipid parameters.
Parameters have been developed and validated for PC~\cite{Chandrasekhar03,kukol09,piggot12},
%  [REFS: https://www.ncbi.nlm.nih.gov/pubmed/12632209, https://pubs.acs.org/doi/10.1021/ct8003468 and https://pubs.acs.org/doi/abs/10.1021/ct3003157],
PE, PG, Cardiolipin~\cite{piggot11}
%[REF: https://pubs.acs.org/doi/abs/10.1021/jp207013v]
and PI~\cite{holdbrook10}
%[REF: https://pubs.acs.org/doi/abs/10.1021/bi101743w]
lipids, amongst others.
Within this work we provide an initial validation for simulating PS lipids with two variants of
this force field (termed GROMOS-CKP and GROMOS-CKPM). Additional forthcoming work will provide additional
validation and testing of these PS parameters. The two variants of the force field only differ in terms
of the charges in the head group. The GROMOS-CKPM (M stands for Mukhopadhyay after the first published
Berger-based PS simulations published; see above for further details) lipid uses Berger/Chiu based charges
for the NH3 part of the PS head group (i.e. has the same charges as used in the choline part of the Berger PC lipids).
The GROMOS-CKP parameters use charges compatible with the rest of the GROMOS force fields;
the  charges of the NH3 part of the PS head group were taken from a lysine side-chain of the GROMOS 54A7 protein
force field~\cite{schmid11} % [REF: https://link.springer.com/article/10.1007%2Fs00249-011-0700-9].
The remainder of the parameters were assigned by analogy from either the other GROMOS-CKP lipids or the GROMOS 54A7 force field.

\noindent
{\it DOPS and POPS bilayers.} 
%The data in Table \ref{PSsystems} and Refs. \citenum{ckp1POPS303K,ckp2POPS303K,ckp1DOPS303K,ckp2DOPS303K}
%\todo{T. Piggot, please write simulation details} \\

Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the CHARMM36 simulations.
A script was used to re-order the atoms and to remove any extraneous hydrogen atoms. The force fields parameters for DOPS and POPS
were as discussed above. The SPC water model was used in these simulations~\cite{berendsen81}. Simulation cut-offs were chosen
to closely match those used in the original GROMOS-CKP force field validation: van der Waals interactions were truncated at 1.4 nm
with a dispersion correction applied to the energy and pressure; Coulombic interactions were also truncated at 1.4 nm with PME used
to treat the long-range interactions. All other simulation settings were identical to those described above for the CHARMM36 DOPS
and POPS simulations bar, including the use of the Verlet cut-off scheme, apart from all bonds were constrained using LINCS in these simulations.

\noindent
{\it POPC:POPS (5:1) mixtures without additional ions}
% The data in Table \ref{mixedIONsystems} and Refs. \citenum{POPCpopsGROMOSCKPwithNa,POPCpopsGROMOSCKPMwithNa}
% \todo{T. Piggot, please write simulation details} \\

Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the equivalent CHARMM36 simulations (see above).
A script was used to re-order the atoms and to remove any extraneous hydrogen atoms. Simulation setting were identical to those as described above
for the GROMOS-CKP(M) POPS bilayers.

@@

\noindent
{\it Thomas Piggot}
Setup, performed and analysed many of the simulations. Contributed to parts of the manuscript.
